\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events

### 2023-09-18 paper/main.tex
finished draft of results section

\newcommand{\stimDescription}{S1} % table
\newcommand{\segments}{S1} % figure
\newcommand{\events}{S2} %
\newcommand{\references}{S3}
\newcommand{\MethodsReplExp}{S4}
\newcommand{\targetAsymmetries}{S5}
\newcommand{\hitRates}{S6}
\newcommand{\characterRefs}{S7}
\newcommand{\refAdjacent}{S8}
\newcommand{\referringReferenced}{S9}

@@

\section*{Introduction}
What we experience in the current moment tells us about \textit{now}-- but what does it tell us about the past or future?  And does the current moment tell us, as human observers, \textit{more} about the past or about the future?  One way of examining these questions is to consider highly simplified scenarios that are artificially constructed in the laboratory \citep[e.g., ][]{MaheEtal22}.  At one extreme, for deterministic sequences with \textit{known} rules, knowing the current state provides the observer with sufficient information to exactly reconstruct the entire past and future history of the stimulus. At another extreme, for purely random sequences, observing the current state provides no information about the past \textit{or} future.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{intro1}
  \caption{\textbf{Retrodiction, retrospection, and prediction.} In one's own life, one may draw on memory to retrospect (i.e., review or re-evaluate) the past or predict the future.  This process is time-asymmetric, since our own past is (typically) observed whereas our future is not.  When we make inferences about \textit{other} people's lives, however, we often have uncertainty about both their past and future, since we may have observed neither.  We may \textit{retrodict} the unobserved past and predict the unobserved future of other people's lives.}
  \label{fig:intro1}
\end{figure}

@@

One key difference between simple artificial sequences and more naturalistic (real or narrative) sequences is that naturalistic sequences often incorporate other people.  Despite the past and future being equally unknown to \textit{the observer} prior to the current moment, other people, and realistic characters in narratives, have their own psychological arrows of time. Specifically, they have memories of their own pasts. Other people's asymmetric knowledge about their \textit{own} pasts and futures might affect their behaviors (e.g., conversations). In turn, this might provide time-asymmetric clues that favor the past~\citep[e.g., other people might talk more about their own pasts than their futures; ][]{DemiEtal18}. If observers leverage these clues from other people's asymmetric knowledge, then observers should also be better at inferring the past (versus the future) of other people's lives.  Alternatively, if inferences about other people's lives are more like inferences about artificial statistical sequences~\citep[e.g., perhaps solely relying on statistical regularities like event schemas, scripts, or situation models][]{RadvCope06, ZwaaRadv98, BoweEtal79, RangRitc12, BaldEtal18}, then the accuracy of inferences about the past and the future of others' lives should be approximately equal.

@@

\section*{Results}
Participants in our main experiment ($n = 36$) watched segments from two storylines, drawn from the CBS television show \textit{Why Women Kill}.  Each storyline comprised 11 segments (mean duration: 2.05~min; range: 0.97--3.87~min, Table~\stimDescription).  We asked participants to use free-form (typed) text responses to retrodict what had happened prior to a just-watched segment, predict what would happen next, or recall what they had just watched (Fig.~\ref{fig:method}, \textit{Task design}). We referred to the to-be-retrodicted, to-be-predicted, or to-be-recalled segment as the \textit{target segment} for each response. We systematically varied whether participants watched the segments in forward or reverse chronological order, and how many segments they had seen prior to making a response (see \textit{Methods}).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{methods}
  \caption{\textbf{Task overview.} Participants in our main experiment watched segments of two storylines from the television series \textit{Why Women Kill}.  They made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment. Across four counterbalanced sequences, we systematically varied whether participants watched the segments in forward or reverse chronological order, whether (or not) responses were cued using the main characters in the target segment, and which other segments participants had watched prior to making a response.  For each segment, we collected several retrodiction, prediction, and/or recall responses across different experimental conditions.  Experiment time is denoted along the
  vertical axis, storyline segments are indicated along the horizontal axis, and the colors denote experimental tasks (conditions).  For an analogous depiction of our replication experiment's design, see Fig.~\MethodsReplExp.}
  \label{fig:method}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.80\textwidth]{results1}
  \caption{\textbf{Retrodiction, prediction, and recall performance by experimental condition.} \textbf{A.  Methods schematic.}  For each retrodiction, prediction, and recall response, we calculated the hit rate for events in the target segment, the response precision (see \textit{Methods}), and the response convergence across participants (see \textit{Methods}).  \textbf{B. Target event hit rate.} Mean proportions of target events that were contained in participants' responses, for each response type, averaged across target segments.  \textbf{C. Response precision.}  Mean precisions of participants' responses, for each response type, averaged across target segments.  The horizontal lines denote the mean pairwise semantic similarities (see \textit{Methods}) across recall responses (re(R): orange; re(P): blue).  \textbf{D. Response convergence.}  Mean (across-participant) convergence of participants' responses, for each response type, averaged across target segments.  All panels: error bars denote bootstrapped 95\% confidence intervals.  Asterisks indicate significance in the (generalized) linear mixed models: * denotes $p < 0.05$ and ** denotes $p < 0.01$.  See Figure~\targetAsymmetries for analogous results from our replication experiment.}
  \label{fig:result1}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{results2}
  \caption{\textbf{Retrodictions and predictions of temporally near and distant events.}  \textbf{A. Illustration of annotation approach.}  For each uncued retrodiction and prediction response in our main experiment, we calculated the number of (retrodicted or predicted) events as a function of temporal distance from the target segment, or \textit{lag}.  Onscreen (explicit) events are tagged using integer-valued lags, whereas offscreen (implicit) events are tagged using half-step lags ($\pm 0.5$, $\pm 1.5$, etc.).  \textbf{B. Number of events hit in participants' uncued retrodictions and predictions for each event type.}  Here we separated events we identified in participants' responses according to whether they occurred in the target segment (lags of $\pm 1$), during the interval between the target segment and the just-watched segment (lags of $\pm 0.5$), at longer temporal distances ($|\mathrm{lag}| > 1$), or were incorrect (unmatched with any past or future events in the narrative).  The counts displayed in the panel are averaged across just-watched segments.  \textbf{C. Number of events hit as a function of temporal distance.}  Here the (across-segment) mean numbers of events hit in participants' uncued retrodictions (orange) and predictions (blue) are displayed as a function of temporal distance to the just-watched segment (lag).  Error bars denote bootstrapped 95\% confidence intervals.  Colors denote temporal direction (orange: past; blue: future) and distance (darker shading: onscreen events from segments adjacent to the target segment; lighter shading: offscreen events).  See Figure~\hitRates for an analogous presentation of results from our replication study.}
  \label{fig:result2}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.9\textwidth]{results3}
  \caption{\textbf{Characters' references drive participants' retrodiction and prediction performance.}  \textbf{A. Illustration of annotation approach.}  We manually annotated references to events in past or future segments in characters' spoken conversations.  We matched each such reference with its corresponding storyline event (and its corresponding segment number for onscreen events, or half-step segment number for offscreen events).  We then tracked the hit rate separately for referenced versus unreferenced events in participants' uncued retrodictions and predictions.  \textbf{B. Reference rate as a function of lag.}  Across all possible just-watched segments (lag 0), the bar heights denote the average proportions of events referenced in other past (orange, negative lags) or future (blue, positive lags) segments in our main experiment's stimulus.  \textbf{C. Difference in hit rates between all events and unreferenced events.}  To highlight the effect of characters' references to past and future events on participants' retrodictions and predictions, here we display the difference in across-segment mean hit rates between all events and unreferenced events, as a function of temporal distance (lag) to the just-watched segment.  \textbf{D. Hit rates for unreferenced events.}  The average response hit rates for unreferenced events are displayed as a function of temporal distance to the just-watched segment.  Error bars denote bootstrapped 95\% confidence intervals.  Panels B--D: colors are described in the Figure~\ref{fig:result2} caption.  \textbf{E.  Hit rates and counts of referenced and unreferenced events.}  As a function of temporal distance to the just-watched segment, the sub-panels display the across-segment mean numbers ($x$-axes) and hit rates ($y$-axes) of referenced (red) and unreferenced (gray) events that participants hit (darker shading) or missed (lighter shading) in their uncued retrodictions (top sub-panel) and uncued predictions (bottom sub-panel).  For an analogous presentation of results from the replication experiment, see Fig.~\characterRefs.}
  \label{fig:result3}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.8\textwidth]{results4}
  \caption{\textbf{Reference-adjacent events are associated with higher hit rates (main experiment).}  \textbf{A. Illustration of annotation approach.}  We extended the annotation procedure depicted in Figure~\ref{fig:result3}A to also label unreferenced events that were either temporally adjacent to (i.e., immediately preceding or proceeding) a referenced event (reference-adjacent events) or not (remaining events).  \textbf{B. Adjacent reference rate for unreferenced events as a function of lag.}  Across all possible just-watched segments (lag 0), the bar heights denote the average proportion of unreferenced events in other past (orange, negative lags) or future (blue, positive lags) segments that were temporally adjacent to any referenced event.  \textbf{C. Difference in hit rates between unreferenced events and remaining events.}  To highlight the effect of reference adjacency on retrodiction and prediction of unreferenced events, here we display the difference in across-segment mean hit rates between unreferenced events and remaining events, as a function of temporal distance (lag) to the just-watched segment.  \textbf{D.  Hit rates for remaining events.}  The across-segment mean response hit rates for unreferenced events that were \textit{not} temporally adjacent to any referenced events are displayed as a function of temporal distance to the just-watched segment. Error bars denote bootstrapped 95\% confidence intervals.  Panels B--D: colors are described in the Figure~\ref{fig:result2} caption.  \textbf{E. Hit rates and counts of referenced, reference-adjacent, and remaining events.}  As a function of temporal distance to the just-watched segment, the sub-panels display the numbers ($x$-axes) and proportions ($y$-axes) of referenced (red), reference-adjacent (purple), and remaining (gray) events that participants hit (darker shading) or missed (lighter shading) in their uncued retrodictions (top sub-panel) and uncued predictions (bottom sub-panel). For an analogous depiction of results from our replication experiment see Fig.~\refAdjacent.}
  \label{fig:result4}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{results5}
  \caption{\textbf{Referenced events are associated with higher hit rates, but referring events are not.}  \textbf{A. Illustration of annotation approach.}  We extended the annotation procedure depicted in Figure~\ref{fig:result3}A to also label which events in our main experiment's stimuli \textit{contained} references to events in other segments.  \textbf{B. Referenced versus referring events.}  During event $i$, when a character makes a reference to another event ($j$), we define $i$ as the \textit{referring} event and $j$ as the \textit{referenced} event.  \textbf{C.  Referring rate as a function of lag.}  Across all possible just-watched segments (lag 0), the bar heights denote the across-segment mean proportions of events containing references to events in other past (orange, negative lags) or future (blue, positive lags) segments.  The bar colors are described in the Figure~\ref{fig:result2} caption.  \textbf{D. Hit rates and counts of referenced, referring, and other events.}  As a function of temporal distance to the just-watched segment, the sub-panels display the numbers ($x$-axes) and hit rates ($y$-axes) of referenced (red), referring (green), and other (gray) events that participants hit (darker shading) or missed (lighter shading) in their uncued retrodictions (top sub-panel) and uncued predictions (bottom sub-panel). For a display of analogous results from our replication experiment see Figure~\referringReferenced.}
  \label{fig:result5}
\end{figure}

@@

\textbf{JRM STOPPED HERE}

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{meta-analysis}
  \caption{\textbf{Meta analysis}}
  \label{fig:meta-analysis}
\end{figure}

\section*{Discussion}
We asked participants to watch sequences of movie segments from a character-driven television drama and then either retrodict what had happened prior to a just-watched segment, predict what would happen next, or recall what they had just watched. We found that participants tended to more accurately and more readily retrodict the unobserved past than predict the unobserved future.  We traced this temporal asymmetry to (a) characters' tendencies to refer to past events more than future events in their ongoing conversations, and (b) associations between temporally proximal events (Fig.~\ref{fig:discussion}). Essentially, associations between temporally proximal events serve to enhance asymmetries in inferences driven by conversational references (light orange and blue bars in Fig.~\ref{fig:discussion}). Our findings show that other peoples' psychological arrows of time can affect external observers' inferences about the unobserved past and future.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{discussion}
  \caption{\textbf{How much information about the past and future can be inferred by observing the present?}  By definition, let us say that the present moment (lag 0) contains all information about itself (dark gray).  Given learned statistical regularities, one might extrapolate from the present moment into the past or future (light gray).  As illustrated in this schematic, the information contained in the present about other moments in time falls off with absolute lag.  This falloff is approximately time-symmetric.  References in the present to past events (dark orange) or future events (dark blue) provide additional information about those referenced moments in time, beyond what could be inferred solely from statistical regularities.  This additional information about those referenced moments can also be extrapolated to other moments that are temporally nearby to \textit{them} (light orange and blue).}
  \label{fig:discussion}
\end{figure}

@@

\section*{Methods}
\subsection*{Participants}
A total of 36 participants (25 female, mean age 21.47 years, range 19--50 years) were recruited from the Dartmouth College community. All participants had self-reported normal or corrected-to-normal vision, hearing, and memory, and had not watched any episodes of \textit{Why Women Kill} before the experiment. Participants gave written consent to enroll in the study under a protocol approved by the Committee for the Protection of Human Subjects at Dartmouth College.  Participants received course credit or monetary compensation for their time. Two participants completed only the first half of the study and one participant’s data from the second half of their testing session was lost due to a technical error. All available data were used in the analyses.

\subsection*{Stimuli}
The stimulus used in the study were segments of the CBS television series \textit{Why Women Kill} Season 1. The TV series contained three distinct storylines depicting three women’s marital relationships. The three storylines, which took place in the 1960s, 1980s, and 2019, were shown in an interleaved fashion in the original episodes. The first 11 segments from the 1960s and 1980s storylines, across the first and second episodes, were used in our study. Segments were divided based on major scene cuts, which primarily corresponded to storyline shifts in the original episodes. The mean length of the segments was 2.05 min (range 0.97--3.87 min). We chose this TV series based on its strictly linear storytelling (within each storyline) and its realistic settings where most events depicted everyday life. The plots were focused on the main characters (Beth in storyline 1 and Simone in storyline 2), who were present in all the segments in the corresponding storylines.

\subsection*{Task design and procedure}
Our experimental paradigm was divided across two testing sessions. In each session, participants performed a sequence of tasks on segments from one storyline (Fig.~\ref{fig:method}).  For each storyline, there were four different task sequences: two forward chronological order sequences and two backward chronological order sequences. Participants completed one task sequence in forward chronological order for one storyline, and one in backward chronological order for the other storyline. The order of the two sessions (forward chronological order sequence first or backward chronological order sequence first), and the pairing of task sequences with storylines, were counterbalanced across participants.

@@

\section*{Code and data availability}
All of the code and data generated for the current manuscript are available online at:

>>>>

% tables
\newcommand{\stimDescription}{S1}
\newcommand{\metaTable}{S2}

% figures
\newcommand{\segments}{S1}
\newcommand{\events}{S2}
\newcommand{\references}{S3}
\newcommand{\MethodsReplExp}{S4}
\newcommand{\targetAsymmetries}{S5}
\newcommand{\hitRates}{S6}
\newcommand{\characterRefs}{S7}
\newcommand{\refAdjacent}{S8}
\newcommand{\referringReferenced}{S9}

@@

\section*{Introduction}

What we experience in the current moment tells us about \textit{now}-- but what
does it tell us about the past or future? And does the current moment tell us,
as human observers, \textit{more} about the past or about the future? One way
of examining these questions is to consider highly simplified scenarios that
are artificially constructed in the laboratory \citep[e.g., ][]{MaheEtal22}. At
one extreme, for deterministic sequences with \textit{known} rules, knowing the
current state provides the observer with sufficient information to exactly
reconstruct the entire past and future history of the stimulus. At another
extreme, for purely random sequences, observing the current state provides no
information about the past \textit{or} future.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{intro1}

  \caption{\textbf{Retrodiction, retrospection, and prediction.} In one's own
  life, one may draw on memory to retrospect (i.e., review or re-evaluate) the
  past or predict the future. This process is time-asymmetric, since our own
  past is (typically) observed whereas our future is not. When we make
  inferences about \textit{other} people's lives, however, we often have
  uncertainty about both their past and future, since we may have observed
  neither. We may \textit{retrodict} the unobserved past and predict the
  unobserved future of other people's lives.}

  \label{fig:intro1}
\end{figure}

@@

One key difference between simple artificial sequences and more naturalistic
(real or narrative) sequences is that naturalistic sequences often incorporate
other people. Despite the past and future being equally unknown to \textit{the
observer} prior to the current moment, other people, and realistic characters
in narratives, have their own psychological arrows of time. Specifically, they
have memories of their own pasts. Other people's asymmetric knowledge about
their \textit{own} pasts and futures might affect their behaviors (e.g.,
conversations). In turn, this might provide time-asymmetric clues that favor
the past~\citep[e.g., other people might talk more about their own pasts than
their futures; ][]{DemiEtal18}. If observers leverage these clues from other
people's asymmetric knowledge, then observers should also be better at
inferring the past (versus the future) of other people's lives. Alternatively,
inferences about other people's lives may be more like inferences about
artificial statistical sequences~\citep[e.g., perhaps solely relying on
statistical regularities like event schemas, scripts, or situation
models][]{RadvCope06, ZwaaRadv98, BoweEtal79, RangRitc12, BaldEtal18}. If so,
then the accuracy of inferences about the past and the future of others' lives
should be approximately equal. We note that the aforementioned authors make no
specific claims about temporal symmetries or asymmetries. Rather, we claim that
statistical regularities might \textit{imply} symmetry (e.g., if you are on
step $n$ of an unfolding schema, this suggests you have just completed step $n
- 1$ and that you are next likely to encounter step $n + 1$).

@@

\section*{Results}

Participants in our main experiment ($n = 36$) watched segments from two
storylines, drawn from the CBS television show \textit{Why Women Kill}. Each
storyline comprised 11 segments (mean duration: 2.05~min; range:
0.97--3.87~min, Table~\stimDescription). We asked participants to use free-form
(typed) text responses to retrodict what had happened prior to a just-watched
segment, predict what would happen next, or recall what they had just watched
(Fig.~\ref{fig:method}, \textit{Task design}). We referred to the
to-be-retrodicted, to-be-predicted, or to-be-recalled segment as the
\textit{target segment} for each response. We systematically varied whether
participants watched the segments in forward or reverse chronological order,
and how many segments they had seen prior to making a response (see
\textit{Methods}).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{methods}

  \caption{\textbf{Task overview.} Participants in our main experiment watched
  segments of two storylines from the television series \textit{Why Women
  Kill}. They made free-form text responses to either retrodict what had
  happened in the previous segment, predict what would happen in the next
  segment, or recall what happened in the just-watched segment. Across four
  counterbalanced sequences, we systematically varied whether participants
  watched the segments in forward or reverse chronological order, whether (or
  not) responses were cued using the main characters in the target segment, and
  which other segments participants had watched prior to making a response. For
  each segment, we collected several retrodiction, prediction, and/or recall
  responses across different experimental conditions. Experiment time is
  denoted along the vertical axis, storyline segments are indicated along the
  horizontal axis, and the colors denote experimental tasks (conditions). For
  an analogous depiction of our replication experiment's design, see
  Fig.~\MethodsReplExp.}

  \label{fig:method}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.80\textwidth]{results1}

  \caption{\textbf{Retrodiction, prediction, and recall performance by
  experimental condition.} \textbf{A. Methods schematic.} For each
  retrodiction, prediction, and recall response, we calculated the hit rate for
  events in the target segment, the response precision (see \textit{Methods}),
  and the response convergence across participants (see \textit{Methods}).
  \textbf{B. Target event hit rate.} Mean proportions of target events that
  were contained in participants' responses, for each response type, averaged
  across target segments. \textbf{C. Response precision.} Mean precisions of
  participants' responses, for each response type, averaged across target
  segments. The horizontal lines denote the mean pairwise semantic similarities
  (see \textit{Methods}) across recall responses (re(R): orange; re(P): blue).
  \textbf{D. Response convergence.} Mean (across-participant) convergence of
  participants' responses, for each response type, averaged across target
  segments. All panels: error bars denote bootstrapped 95\% confidence
  intervals. Asterisks indicate significance in the (generalized) linear mixed
  models: * denotes $p < 0.05$ and ** denotes $p < 0.01$. See
  Figure~\targetAsymmetries for analogous results from our replication
  experiment.}

  \label{fig:result1}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{results2}

  \caption{\textbf{Retrodictions and predictions of temporally near and distant
  events.} \textbf{A. Illustration of annotation approach.} For each uncued
  retrodiction and prediction response in our main experiment, we calculated
  the number of (retrodicted or predicted) events as a function of temporal
  distance from the target segment, or \textit{lag}. Onscreen (explicit) events
  are tagged using integer-valued lags, whereas offscreen (implicit) events are
  tagged using half-step lags ($\pm 0.5$, $\pm 1.5$, etc.). \textbf{B. Number
  of events hit in participants' uncued retrodictions and predictions for each
  event type.} Here we separated events we identified in participants'
  responses according to whether they occurred in the target segment (lags of
  $\pm 1$), during the interval between the target segment and the just-watched
  segment (lags of $\pm 0.5$), at longer temporal distances ($|\mathrm{lag}| >
  1$), or were incorrect (unmatched with any past or future events in the
  narrative). The counts displayed in the panel are averaged across
  just-watched segments. \textbf{C. Number of events hit as a function of
  temporal distance.} Here the (across-segment) mean numbers of events hit in
  participants' uncued retrodictions (orange) and predictions (blue) are
  displayed as a function of temporal distance to the just-watched segment
  (lag). Error bars denote bootstrapped 95\% confidence intervals. Colors
  denote temporal direction (orange: past; blue: future) and distance (darker
  shading: onscreen events from segments adjacent to the target segment;
  lighter shading: offscreen events). See Figure~\hitRates for an analogous
  presentation of results from our replication study.}

  \label{fig:result2}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.9\textwidth]{results3}

  \caption{\textbf{Characters' references drive participants' retrodiction and
  prediction performance.} \textbf{A. Illustration of annotation approach.} We
  manually annotated references to events in past or future segments in
  characters' spoken conversations. We matched each such reference with its
  corresponding storyline event (and its corresponding segment number for
  onscreen events, or half-step segment number for offscreen events). We then
  tracked the hit rate separately for referenced versus unreferenced events in
  participants' uncued retrodictions and predictions. \textbf{B. Reference rate
  as a function of lag.} Across all possible just-watched segments (lag 0), the
  bar heights denote the average proportions of events referenced in other past
  (orange, negative lags) or future (blue, positive lags) segments in our main
  experiment's stimulus. \textbf{C. Difference in hit rates between all events
  and unreferenced events.} To highlight the effect of characters' references
  to past and future events on participants' retrodictions and predictions,
  here we display the difference in across-segment mean hit rates between all
  events and unreferenced events, as a function of temporal distance (lag) to
  the just-watched segment. \textbf{D. Hit rates for unreferenced events.} The
  average response hit rates for unreferenced events are displayed as a
  function of temporal distance to the just-watched segment. Error bars denote
  bootstrapped 95\% confidence intervals. Panels B--D: colors are described in
  the Figure~\ref{fig:result2} caption. \textbf{E. Hit rates and counts of
  referenced and unreferenced events.} As a function of temporal distance to
  the just-watched segment, the sub-panels display the across-segment mean
  numbers ($x$-axes) and hit rates ($y$-axes) of referenced (red) and
  unreferenced (gray) events that participants hit (darker shading) or missed
  (lighter shading) in their uncued retrodictions (top sub-panel) and uncued
  predictions (bottom sub-panel). For an analogous presentation of results from
  the replication experiment, see Fig.~\characterRefs.}

  \label{fig:result3}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.8\textwidth]{results4}

  \caption{\textbf{Reference-adjacent events are associated with higher hit
  rates (main experiment).} \textbf{A. Illustration of annotation approach.} We
  extended the annotation procedure depicted in Figure~\ref{fig:result3}A to
  also label unreferenced events that were either temporally adjacent to (i.e.,
  immediately preceding or proceeding) a referenced event (reference-adjacent
  events) or not (remaining events). \textbf{B. Adjacent reference rate for
  unreferenced events as a function of lag.} Across all possible just-watched
  segments (lag 0), the bar heights denote the average proportion of
  unreferenced events in other past (orange, negative lags) or future (blue,
  positive lags) segments that were temporally adjacent to any referenced
  event. \textbf{C. Difference in hit rates between unreferenced events and
  remaining events.} To highlight the effect of reference adjacency on
  retrodiction and prediction of unreferenced events, here we display the
  difference in across-segment mean hit rates between unreferenced events and
  remaining events, as a function of temporal distance (lag) to the
  just-watched segment. \textbf{D. Hit rates for remaining events.} The
  across-segment mean response hit rates for unreferenced events that were
  \textit{not} temporally adjacent to any referenced events are displayed as a
  function of temporal distance to the just-watched segment. Error bars denote
  bootstrapped 95\% confidence intervals. Panels B--D: colors are described in
  the Figure~\ref{fig:result2} caption. \textbf{E. Hit rates and counts of
  referenced, reference-adjacent, and remaining events.} As a function of
  temporal distance to the just-watched segment, the sub-panels display the
  numbers ($x$-axes) and proportions ($y$-axes) of referenced (red),
  reference-adjacent (purple), and remaining (gray) events that participants
  hit (darker shading) or missed (lighter shading) in their uncued
  retrodictions (top sub-panel) and uncued predictions (bottom sub-panel). For
  an analogous depiction of results from our replication experiment see
  Fig.~\refAdjacent.}

  \label{fig:result4}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{results5}

  \caption{\textbf{Referenced events are associated with higher hit rates, but
  referring events are not.} \textbf{A. Illustration of annotation approach.}
  We extended the annotation procedure depicted in Figure~\ref{fig:result3}A to
  also label which events in our main experiment's stimuli \textit{contained}
  references to events in other segments. \textbf{B. Referenced versus
  referring events.} During event $i$, when a character makes a reference to
  another event ($j$), we define $i$ as the \textit{referring} event and $j$ as
  the \textit{referenced} event. \textbf{C. Referring rate as a function of
  lag.} Across all possible just-watched segments (lag 0), the bar heights
  denote the across-segment mean proportions of events containing references to
  events in other past (orange, negative lags) or future (blue, positive lags)
  segments. The bar colors are described in the Figure~\ref{fig:result2}
  caption. \textbf{D. Hit rates and counts of referenced, referring, and other
  events.} As a function of temporal distance to the just-watched segment, the
  sub-panels display the numbers ($x$-axes) and hit rates ($y$-axes) of
  referenced (red), referring (green), and other (gray) events that
  participants hit (darker shading) or missed (lighter shading) in their uncued
  retrodictions (top sub-panel) and uncued predictions (bottom sub-panel). For
  a display of analogous results from our replication experiment see
  Figure~\referringReferenced.}

  \label{fig:result5}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{meta-analysis}

\caption{\textbf{Meta analysis.} We used natural language processing to
automatically identify references to past or future events across a variety of
sources. \textbf{A. Numbers of past and future events in \textit{The Chair},
Season 1, Episodes 1--6.} The bar heights indicate the raw numbers of manually
identified (lighter shading) and automatically identified (darker shading) past
and future events from each episode (color). We used Episode 1 from this series
as the stimulus in our replication experiment. \textbf{B. Proportions of past
and future events in \textit{The Chair}, Season 1, Episodes 1--6.} The Panel is
in the same format as Panel A, but here the bar heights have been divided by
the total numbers of past and future events (per episode). \textbf{C.
Proportions of past and future events in movies, television shows, and natural
conversations.} As in Panel B, the bar heights denote the proportions of past
and future events detected in each dataset (color). The datasets are described
in Table~\metaTable. Error bars denote bootstrap-estimated 95\% confidence
intervals.}

  \label{fig:meta-analysis}
\end{figure}

The above analyses show that characters in the television shows we used as
stimuli in our main experiment and replication experiment refer more often to
the past than to the future. This appears to bias participants' inferences
about the past and future. But how universal is this pattern? For example, were
the television shows we happened to select for our experiment representative of
television shows more generally? Or perhaps media created for entertainment
purposes tends to have a bias towards the past in order to keep the story
engaging and unpredictable. To better understand temporal biases in
conversations, we carried out a meta analysis using extracted conversation data
from several large datasets, comprising a total of over 440 million
conversations from over 17 million documents. The data comprised transcripts
from television shows and popular films, novels, and spoken and written
utterances from natural conversations. A summary of the data we analyzed may be
found in Table~\metaTable. As summarized in Figure~\ref{fig:meta-analysis}, we
used natural language processing to identify references to past or future
events in each conversation (also see \textit{Meta analysis of conversation
data}).

To validate our basic approach, we compared the numbers
(Fig.~\ref{fig:meta-analysis}A) and proportions (Fig.~\ref{fig:meta-analysis}B)
of automatically and manually identified references to past and future events,
across six episodes of the television show \textit{The Chair}. (The first
episode was used as the stimulus in our replication study.) In general, our
automated tagging procedure tended to overcount the numbers of references. From
manually inspecting hundreds of example tags, we believe this discrepancy
follows from which criteria were used to generate the tags. The manually
generated tags sought to identify references to specific events that occurred
or were implied to occur in other parts of the narrative. In contrast, as a
heuristic, we designed the automatic tagging procedure to identify uses of the
past or future \textit{tense} as a proxy for references to past or future
\textit{events}. We noticed that, a single conversation often contains multiple
references to a given (past or future) event. Whereas the manually generated
tags counted these as ``single'' references, our automated tagging procedure
had no means of differentiating between multiple references to the same event
versus references to different events. Nevertheless, this discrepancy did not
appear to bias the balance of the overall \textit{proportions} of past or
future references.

In all, across all of the datasets we examined in our meta analysis, we
identified a total of 36,008,500 references to past or future events. A total
of 19,464,741 (54.06\%) of these were references to past events, and the
remaining 16,543,759 (45.94\%) were references to future events. We also
computed the average proportions of references to past and future events across
documents within each individual dataset. Across the 12 datasets we examined
(Fig.~\ref{fig:meta-analysis}, Tab.~\metaTable), there were significantly more
references to the past than the to the future (mean $\pm$ standard deviation:
$58.99\% \pm 7.28\%$; $t(11) = 4.28, p = 0.0013$). This bias towards the past
also held for each dataset individually ($t$s $\geq 5.14$, $p$s $< 0.01$)
except for one dataset, ``Persuasion for Good,'' which comprised natural
conversations between pairs of Amazon Mechanical Turk workers wherein one
participant tried to convince the other participant to donate to a charity in
the future. In that dataset, references to the future were significantly more
common than references to the past ($t(11438) = -22.65, p < 0.001$). This
latter example provided a nice sanity check for verifying that our general
approach was not itself biased in favor of the past, e.g., even in
conversations that were actually biased towards the future. Taken together, the
results from our meta analysis indicate that people tend to refer to the past
more than they refer to the future, across a wide variety of situations
(including in both fictional and real conversations). Although (as in the
Persuasion for Good dataset) there may be specific exceptions to this bias, it
seems that a bias in favor of the past is a common element of many (and perhaps
even \textit{most}) human conversations.

\section*{Discussion}

We asked participants to watch sequences of movie segments from a
character-driven television drama and then either retrodict what had happened
prior to a just-watched segment, predict what would happen next, or recall what
they had just watched. We found that participants tended to more accurately and
more readily retrodict the unobserved past than predict the unobserved future.
We traced this temporal asymmetry to (a) characters' tendencies to refer to
past events more than future events in their ongoing conversations, and (b)
associations between temporally proximal events (Fig.~\ref{fig:discussion}).
Essentially, associations between temporally proximal events serve to enhance
asymmetries in inferences driven by conversational references (light orange and
blue bars in Fig.~\ref{fig:discussion}). Our findings show that other peoples'
psychological arrows of time can affect external observers' inferences about
the unobserved past and future.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{discussion}

  \caption{\textbf{How much information about the past and future can be
  inferred by observing the present?} By definition, let us say that the
  present moment (lag 0) contains all information about itself (dark gray).
  Given learned statistical regularities, one might extrapolate from the
  present moment into the past or future (light gray). As illustrated in this
  schematic, the information contained in the present about other moments in
  time falls off with absolute lag. This falloff is approximately
  time-symmetric. References in the present to past events (dark orange) or
  future events (dark blue) provide additional information about those
  referenced moments in time, beyond what could be inferred solely from
  statistical regularities. This additional information about those referenced
  moments can also be extrapolated to other moments that are temporally nearby
  to \textit{them} (light orange and blue).}

  \label{fig:discussion}
\end{figure}

@@

\section*{Methods}
\subsection*{Participants}

A total of 36 participants (25 female, mean age 21.47 years, range 19--50
years) were recruited from the Dartmouth College community. All participants
had self-reported normal or corrected-to-normal vision, hearing, and memory,
and had not watched any episodes of \textit{Why Women Kill} before the
experiment. Participants gave written consent to enroll in the study under a
protocol approved by the Committee for the Protection of Human Subjects at
Dartmouth College. Participants received course credit or monetary compensation
for their time. Two participants completed only the first half of the study and
one participant’s data from the second half of their testing session was lost
due to a technical error. All available data were used in the analyses.

\subsection*{Stimuli}

The stimulus used in the study were segments of the CBS television series
\textit{Why Women Kill} Season 1. The TV series contained three distinct
storylines depicting three women’s marital relationships. The three storylines,
which took place in the 1960s, 1980s, and 2019, were shown in an interleaved
fashion in the original episodes. The first 11 segments from the 1960s and
1980s storylines, across the first and second episodes, were used in our study.
Segments were divided based on major scene cuts, which primarily corresponded
to storyline shifts in the original episodes. The mean length of the segments
was 2.05 min (range 0.97--3.87 min). We chose this TV series based on its
strictly linear storytelling (within each storyline) and its realistic settings
where most events depicted everyday life. The plots were focused on the main
characters (Beth in storyline 1 and Simone in storyline 2), who were present in
all the segments in the corresponding storylines.

\subsection*{Task design and procedure}

Our experimental paradigm was divided across two testing sessions. In each
session, participants performed a sequence of tasks on segments from one
storyline (Fig.~\ref{fig:method}). For each storyline, there were four
different task sequences: two forward chronological order sequences and two
backward chronological order sequences. Participants completed one task
sequence in forward chronological order for one storyline, and one in backward
chronological order for the other storyline. The order of the two sessions
(forward chronological order sequence first or backward chronological order
sequence first), and the pairing of task sequences with storylines, were
counterbalanced across participants.

@@

\section*{Code and data availability}

All of the code and data generated for the current manuscript are available
online at:
